---
title: "Cross-Validation Methods: Ridge Regression"
date: today
---

```{r}
#| echo: false
source("../assets/notes-setup.R")
```


Recall when we had used the data in **equal-educational-opportunity.csv** to examine whether the level of school facilities was an important predictor of student achievement after accounting for the variation in faculty credentials and peer influence, the model suffered from strong collinearity resulting in poor estimates of the coefficients and very large SEs. This means we could not effectively control for peer influence and faculty credentials, which means we could not answer the RQ.

One method of dealing with collinearity we learned about was to use a biased estimation method. The biased estimation method introduced was ridge regression. One issue with fitting a ridge regression model is choosing the ridge penalty, or $\lambda$ value. One good way of choosing $\lambda$ is via cross-validation. The algorithm for choosing $\lambda$ via 10-fold cross-validation is as follows:

- Split the dataset randomly into 10 groups
- For each unique group:
    + Hold the group out as test data set
    + Take the remaining 9 groups and use them as a training data set
    + Fit a ridge model on the training set with a specific value of $\lambda$
    + Evaluate the model  using the test set by computing the amount of error (e.g., MSE, deviance)
    + Repeat for different values of $\lambda$
- Summarize the performance for the various $\lambda$ values by averaging across the 10 different tests sets
- Choose the $\lambda$ value that gives the lowest average amount of error




```{r}
# Load libraries
library(MASS)
library(modelr)
library(tidymodels)
library(tidyverse)

# Read in data
eeo = read_csv("https://raw.githubusercontent.com/zief0002/symmetrical-fiesta/main/data/equal-education-opportunity.csv")

# Standardize all variables in the eeo data frame
z_eeo = eeo |>
  scale() |>
  data.frame()

# View data
head(z_eeo)
```

<br />


## 10-Fold Cross Validation for Ridge Regression

Rather than use the `crossv_kfold()` function from the `{modelr}` package to create the training and validation data sets, we will do this manually. Since there are 70 cases in the data, each of the 10 folds will include 7 cases.


```{r}
# Create a vector that includes 7 cases in each of the 10 folds
fold = rep(1:10, 7)

# Randomly assign each case to a fold
set.seed(100) #Set seed for reproducible results
z_eeo = z_eeo |>
  mutate(
    fold = sample(fold)
  )

# View data
head(z_eeo)
```


Next, we will write a function that will carry out the cross-validation by inputting both a training and validation dataset, fit the ridge regression using different $\lambda$ values to the training data, and the use those coefficients to compute residuals and the MSE for the validation data.



```{r}
ridge_mse = function(.training, .validation){
  
  # Fit ridge model across several lambda values to the training data
  ridge_models = lm.ridge(achievement ~ 1 + faculty + peer + school, 
                          data = .training,
                          lambda = seq(from = 0, to = 100, by = 0.01)
                          )
  
  # Use coefficients from ridge_models with validation data
  # Get design matrix and outcome
  X = as.matrix(.validation[c("faculty", "peer", "school")])
  y = .validation$achievement
  
  # How many lambda values?
  n_lambda = length(ridge_models$lambda)
  
  # Create empty vector to store MSE values
  MSE = c()
  
  # Loop through different sets of coefficients to compute MSE
  for(i in 1:n_lambda){
    b = matrix(ridge_models$coef[ , i]) #Get coefficients for a given lambda
    e = y - (X %*% b) #Compute residual vector
    MSE[i] = sum(e ^ 2) / length(e) #Compute MSE
  }
  
  # Create data frame of lambda and associated MSE values
  d = data.frame(
    lambda = ridge_models$lambda,
    MSE = MSE
  )
  
  return(d)
}
```

Now we will test our function by using fold 1 as the validation data and folds 2--10 in the training data


```{r}
#| eval: false

# Create datasets
train = z_eeo |> filter(fold != 1)
validate = z_eeo |> filter(fold == 1)

# Test function
ridge_mse(.training = train, .validation = validate)
```


Since the function seems to work (not shown), we will write a FOR loop to cycle through each of the 10 different ways we can split the 10 folds into 9 training folds and 1 validation fold.

```{r}
# Create an empty list to store dataframes of the results for each of the 10 cross-validations
my_results = list()

for(i in 1:10){
  
  # Create datasets
  train = z_eeo |> filter(fold != i)
  validate = z_eeo |> filter(fold == i)

  # Carry out cross-validation
  my_results[[i]] = ridge_mse(.training = train, .validation = validate) |>
    mutate(
      v_fold = i
    )
}

# Stack the 10 datframes into a single dataframe
cv_results = do.call(rbind, my_results)
```

Now want to average the MSE for each $\lambda$ value.

```{r}
cv_results |>
  group_by(lambda) |>
  summarize(
    CV_MSE = mean (MSE)
  ) |>
  arrange(CV_MSE) |>
  data.frame() |>
  head(10) #only show the first 10 rows
```


This suggests that a $\lambda$ value of 12.90 may be optimal.

```{r}
# Fit model with best lambda value
best_ridge = lm.ridge(achievement ~ 1 + faculty + peer + school, 
                          data = z_eeo,
                          lambda = 12.90
                          )

# View coefficients
tidy(best_ridge)
```

Based on $\lambda=12.90$ value chosen from the cross-validation, the fitted equation is:

$$
\hat{\mathrm{Achievement}}_i = 0.114(\mathrm{Faculty}_i) + 0.206(\mathrm{Peer}_i) + 0.085(\mathrm{School}_i)
$$
where all of the variables are standardized.

<br />


## Using the `glmnet()` Function to Select $\lambda$


The `cv.glmnet()` function from the `{glmnet}` packages uses *k*-fold cross-validation to select a $\lambda$ value without the hassle of writing the functions and for loops. This function takes the following arguments: 

- `x=` the design matrix
- `y=` the vector of the outcome
- `nfolds=` is the number of folds to use in the CV
- `type.measure=` sets the metric to compute (e.g., "mse", "deviance", "mae")
- `alpha=0` This tells the function to fit a ridge regression
- `intercept=` This is either TRUE or FALSE depending on whether you want it to include an intercept in the model. (With a standardized model, set this to FALSE.)
- `standardize=` This is also either TRUE or FALSE. (If you have pre-standardized X and Y, then set this to FALSE.)



<!-- To use cross-validation for *d* selection, we perform the cross-validation using the `cv.glmnet()` function. Then, we can plot the results and extract the "best" *d* value. -->

```{r}
# Load library
library(glmnet)

# Get matrices for use in glmnet
X = scale(eeo)[ , 2:4]
Y = scale(eeo)[ , 1]

# Make results reproducible
set.seed(100)

# Carry out the cross-validation
ridge_cv = cv.glmnet(
  x = X,
  y = Y,
  nfolds = 10,
  type.measure = "mse",
  alpha = 0,
  intercept = FALSE,
  standardize = FALSE
  )

# Extract the lambda value with the lowest mean error
ridge_cv$lambda.min
```


Here the $\lambda$ value with the lowest mean error from the cross-validation results is `r ridge_cv$lambda.min`. 


:::fyi
**FYI**

The functions in `{glmnet}` optimize using the *penalized mean square error*, whereas the `lm.ridge()` function from `{MASS}` optimizes using the *penalized least squares*. Because of this the $\lambda$ value from `{glmnet}` are scaled by the sample size. To get the $\lambda$ value for use in either the `lm.ridge()` function or directly in the matrix algebra, we need to multiply the `{glmnet}` $\lambda$ value by the sample size:

$$
\lambda_{\mathrm{lm.ridge}} = \lambda_{\mathrm{glmnet}} \times n
$$

In our example this is $0.2108243 \times 70 = 14.7577$.
:::


Once we have the $\lambda$ value from the cross-validation, we can use this directly in the `glmnet()` function to obtain the ridge regression coefficients.


```{r}
# Fit ridge regression
ridge_best = glmnet(
  x = X,
  y = Y,
  alpha = 0,
  lambda = ridge_cv$lambda.min,
  intercept = FALSE,
  standardize = FALSE
  )

# Show coefficients
tidy(ridge_best)
```

Based on $\lambda=0.2108243$ value chosen from the cross-validation, the fitted equation is:

$$
\hat{\mathrm{Achievement}}_i = 0.116(\mathrm{Faculty}_i) + 0.197(\mathrm{Peer}_i) + 0.091(\mathrm{School}_i)
$$

where all of the variables are standardized.

<br />




<!-- ## Least Absolute Shrinkage and Selection Operator (LASSO) -->

<!-- One disadvantage to ridge regression is that although the coefficients are "shrunk", they are not set to 0 (unless $d=\infty$). In some cases this is fine, but in many applications it may be of interest to find the subset of correlated predictors that are useful. This is especially true when your data have a large number of predictors and a limited number of cases (the so-called "large *p*, small *n* problems). In such cases, the LASSO [@Tibshirani:1996] is an alternative to ridge regression. Similar to ridge regression, the LASSO finds the coefficients that minimize a penalized sum of squares, namely: -->

<!-- $$ -->
<!-- \mathrm{SSE}_{\mathrm{Penalized}} = \sum_{i=1}^n \bigg(y_i - \hat y_i\bigg)^2 + d \sum_{j=1}^p \vert \beta_j \vert -->
<!-- $$ -->

<!-- If we compare this to ridge regression, the only difference is in the penalty term. Namely, the penalty term for ridge regression was based on the sum of the squared beta terms while that for the LASSO is based on the sum of the absolute value for the beta terms. -->

<!-- :::fyi -->
<!-- In statistical nomenclature, the LASSO uses an L1 penalty and ridge regression uses an L2 penalty. This terminology is based on the norm (magnitude) of the vector of beta coefficients. The L1 norm is the sum of the absolute values (Manhattan distance) and the L2 norm is the sum of the squared values (Euclidean distance). -->
<!-- ::: -->

<!-- Similar to ridge regression, the penalty term has the effect of biasing some coefficients by shrinking them toward zero. The LASSO, however, will shrink some coefficients to 0; effectively removing them from the model. We can also use the `glmnet()` function from the **glmnet** package to fit a LASSO. To fit a ridge regression we provide the following arguments to `glmnet()`: -->

<!-- - `x=`: A matrix of the predictors -->
<!-- - `y=`: A vector of the outcome -->
<!-- - `alpha=1`: This fits a LASSO -->
<!-- - `lambda=`: This sets a modified $d$ value which is equal to $\frac{d}{n}$ -->
<!-- - `intercept=FALSE`: We will include this argument to drop the intercept term; appropriate for standardized regression. -->
<!-- - `standardize=`: Here we set this to `FALSE` since we pre-standardized the predictors and outcome. The default is `standardize=TRUE`. -->

<!-- <br /> -->

<!-- ### Choosing the LASSO Penalty -->

<!-- Again, we have to choose the modified *d* value to put it the `glmnet()` function. To select *d*, we will again use cross-validation. -->




<!-- ```{r} -->
<!-- # Make reproducible -->
<!-- set.seed(100) -->

<!-- # Fit cross-validated LASSO -->
<!-- lasso_1 = cv.glmnet( -->
<!--   x = X, -->
<!--   y = Y, -->
<!--   alpha = 1, -->
<!--   intercept = FALSE, -->
<!--   standardize = FALSE -->
<!--   ) -->

<!-- # Extract d value with lowest CV-MSE -->
<!-- lasso_1$lambda.min -->
<!-- ``` -->

<!-- Using this value in the LASSO: -->


<!-- ```{r} -->
<!-- # Fit LASSO -->
<!-- best_lasso = glmnet( -->
<!--   x = X, -->
<!--   y = Y, -->
<!--   alpha = 1, -->
<!--   lambda = lasso_1$lambda.min, -->
<!--   intercept = FALSE, -->
<!--   standardize = FALSE -->
<!--   ) -->

<!-- # Show coefficients -->
<!-- tidy(best_lasso) -->
<!-- ``` -->

<!-- Based on using the modified *d* value of 0.042, the fitted LASSO model is: -->

<!-- $$ -->
<!-- \hat{\mathrm{Achievement}}_i = 0.397(\mathrm{Peer}_i) -->
<!-- $$ -->

<!-- where the variables are all standardized. -->


<!-- Note that the coefficients for faculty credentials and school facilities were both shrunk to zero! Computing standard errors for LASSO coefficients are problematic in the same way that computing SEs are problematic for the ridge regression; namely we have no good estimate of the true amount of bias. Lastly, we note that, as in ridge regression, selecting a good value of *d* for the LASSO is critical.  -->

<!-- Although better than ridge regression, the LASSO is not a panacea for model selection under collinearity. As we saw in our example, when the predictors are highly correlated, the LASSO will tend to select a single variable from the set of predictors and ignore the others. It also will only choose at most *n* variables when there are a large number of predictors and small number of cases (large *p*; small *n*). -->



<!-- <br /> -->

<!-- ## Elastic Net -->

<!-- A generalization of the ridge and lasso penalties, introduced by @Zou:2005 is called the *elastic net*. This combines the two penalties and subsequently minimizes a double-penalized sum of squares, namely: -->

<!-- $$ -->
<!-- \mathrm{SSE}_{\mathrm{Penalized}} = \sum_{i=1}^n \bigg(y_i - \hat y_i\bigg)^2 + \bigg(d_1 \sum^p_{j=1} \beta_j^2 + d_2 \sum^p_{j=1} \vert \beta_j \vert \bigg) -->
<!-- $$ -->

<!-- The elastic net has two penalty parameters, namely $d_1$ and $d_2$. As can be seen by this equation, the ridge regression and LASSO are special cases of the elastic net. This reduces to the ridge regression model when $d_2=0$ and to the LASSO when $d_1=0$. When both penalty parameters are non-zero, this results in a mixture of both the L1 and L2 penalties. -->

<!-- Recall that this is simply, -->

<!-- $$ -->
<!-- \mathrm{SSE}_{\mathrm{Penalized}} = \mathrm{SSE} + \mathrm{Penalty} -->
<!-- $$ -->

<!-- In the `glmnet()` function we set an argument `alpha=`. This is actually a mixture parameter that defines the penalty term: -->

<!-- $$ -->
<!-- \mathrm{Penalty} = \frac{1-\alpha}{2}\bigg(\beta_j^2\bigg) + \alpha\bigg(\vert \beta_j \vert\bigg) -->
<!-- $$ -->

<!-- When $\alpha=0$ the penalty is a scaled version of the L1 penalty, $\beta_j^2$; namely ridge regression. (Dividing the ridge penalty by 2 is a mathematical convenience for optimization.) When $\alpha=1$ the penalty is the L2 penalty; namely LASSO. To include both the L1 and L2 penalties we set `alpha=0.5`.  -->



<!-- ```{r} -->
<!-- # Make reproducible -->
<!-- set.seed(1000000) -->

<!-- # Fit cross-validated elastic net  -->
<!-- elastic_net_1 = cv.glmnet( -->
<!--   x = X, -->
<!--   y = Y, -->
<!--   alpha = 0.5, -->
<!--   intercept = FALSE, -->
<!--   standardize = FALSE -->
<!--   ) -->

<!-- # Extract d value with lowest CV-MSE -->
<!-- elastic_net_1$lambda.min -->
<!-- ``` -->

<!-- We re-fit the elastic net using this modified *d* value in the `lambda=` argument. -->

<!-- ```{r} -->
<!-- # Fit elastic net -->
<!-- best_elastic_net = glmnet( -->
<!--   x = X, -->
<!--   y = Y, -->
<!--   alpha = 0.5, -->
<!--   lambda = elastic_net_1$lambda.min, -->
<!--   intercept = FALSE, -->
<!--   standardize = FALSE -->
<!--   ) -->

<!-- # Get coefficients -->
<!-- tidy(best_elastic_net) -->
<!-- ``` -->

<!-- the fitted elastic net model is: -->

<!-- $$ -->
<!-- \hat{\mathrm{Achievement}}_i = 0.391(\mathrm{Peer}_i) + 0.028(\mathrm{Faculty}_i) -->
<!-- $$ -->

<!-- where the variables are all standardized. Here the model still finds a moderate positive effect of peer influence, but there is also a small effect of faculty. Although shrunken toward zero relative to the ridge regression model, it is not 0 as in the LASSO.  -->

<!-- <br /> -->

<!-- ## Non-Collinear Predictors -->

<!-- These methods can also be used when the predictors are not collinear. The elastic net or LASSO is a good option in this case as both essentially do automatic variable selection (by shrinking some of the coefficients to zero) and perform coefficient estimation, simultaneaously. Here we use the elastic net, as the degree of shrinkage is less than the LASSO, to select a model for our life expectancy model using the *states-2019.csv* data. -->

<!-- ```{r} -->
<!-- usa = read_csv("../data/states-2019.csv") -->

<!-- X = usa[ , 3:9] %>% scale() -->
<!-- Y = usa[ , 2] %>% scale() -->

<!-- # Make reproducible -->
<!-- set.seed(3) -->

<!-- # Fit cross-validated elastic net  -->
<!-- en_results = cv.glmnet( -->
<!--   x = X, -->
<!--   y = Y, -->
<!--   alpha = 0.5, -->
<!--   intercept = FALSE, -->
<!--   standardize = FALSE -->
<!--   ) -->

<!-- # Extract d value with lowest CV-MSE -->
<!-- en_results$lambda.min -->

<!-- # Re-fit elastic net -->
<!-- en_model = glmnet( -->
<!--   x = X, -->
<!--   y = Y, -->
<!--   alpha = 0.5, -->
<!--   lambda = en_results$lambda.min, -->
<!--   intercept = FALSE, -->
<!--   standardize = FALSE -->
<!--   ) -->

<!-- # Get coefficients -->
<!-- tidy(en_model) -->
<!-- ``` -->

<!-- The model slected is: -->

<!-- $$ -->
<!-- \hat{\mathrm{Life~Expectancy}}_i = 0.195(\mathrm{Income}_i) - 0.042(\mathrm{Murder~Rate}_i) -->
<!-- $$ -->

<!-- The model selected the predictors of income, and murder rate. The problem with using biased regression methods when there is not a collinearity issue is that it unnecessarily introduces bias into the coefficient estimates---the effects are smaller than when estimated with OLS. So, while it selected a small subset of predictors, we may not have reasonable estimates of the effects. It is not recommended to use shrinkage methods when there is little/no collinearity problems. -->


<!-- <br /> -->

<!-- ## References -->

